---
title: "Code-Through Tutorial"
subtitle: "Introduction to Data Visualization in R using ggplot2"
author: "Lara Wehbe"
output:
  ioslides_presentation
---

# 1. Introduction

## Overview

In this tutorial, we will explore the basics of data visualization in R using the `ggplot2` package. Data visualization is a crucial aspect of data analysis, allowing us to understand patterns, trends, and insights effectively. `ggplot2` is a powerful and flexible package for creating static and interactive plots in R.

## Why Use `ggplot2`?

- Easy-to-use layered grammar of graphics
- Highly customizable
- Can handle large datasets efficiently
- Supports various types of plots

# 2. Setting Up the Environment

## Load the needed libraries

```{r, message=FALSE, warning=FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(maps)
library(ggthemes)
library(kableExtra) # Added for better table rendering
```

# 3. Exploring a Dataset

## Overview

We will use the built-in `mtcars` dataset, which contains data on different car models, including miles per gallon (mpg), horsepower (hp), weight (wt), and other characteristics.

```{r}
# View first few rows of the dataset
head(mtcars)
```

## Summary Statistics

```{r, results='asis', echo=FALSE}
# Summary statistics of the dataset
knitr::kable(summary(mtcars), format = "html") %>% 
  kable_styling(full_width = FALSE, position = "center", font_size = 12) %>%
  scroll_box(height = "400px") # Enables scrolling for better visibility
```

# 4. Data Visualization

## Scatter Plot: MPG vs Horsepower

A scatter plot is useful for visualizing relationships between two continuous variables. Here, we plot `mpg` against `hp`.

```{r, echo=FALSE}
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point(color = "blue") +
  labs(title = "Scatter Plot of MPG vs Horsepower",
       x = "Horsepower",
       y = "Miles Per Gallon")
```

## Histogram: Distribution of MPG

A histogram helps visualize the distribution of a single numeric variable. Below, we create a histogram of `mpg`.

```{r, echo=FALSE}
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Miles Per Gallon",
       x = "Miles Per Gallon",
       y = "Count")
```

# 5. Map of Fuel Efficiency by State

## Introduction

A choropleth map visualizes data using color shading across geographical regions. Here, we create a state-level map of average miles per gallon (mpg) using synthetic data.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
# Load state map data
us_states <- map_data("state")

# Generate synthetic average mpg data for each state
set.seed(42)
state_mpg <- data.frame(
  region = tolower(unique(us_states$region)),
  avg_mpg = runif(length(unique(us_states$region)), min = 15, max = 30)
)

# Merge with map data
map_data_mpg <- merge(us_states, state_mpg, by = "region", all.x = TRUE)

# Plot the choropleth map
MapFig <- ggplot(map_data_mpg, aes(long, lat, group = group, fill = avg_mpg)) + 
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  theme_minimal() +
  scale_fill_distiller(palette = "Blues", direction = 1, name = "Avg MPG") +
  labs(title = "Average Fuel Efficiency by State",
       caption = "Synthetic Data for Demonstration Purposes")

MapFig
```

# 6. Conclusion

## Summary

- `ggplot2` is a powerful tool for data visualization in R.
- We explored scatter plots, histograms, and a choropleth map for fuel efficiency.
- Customization options allow us to create informative and visually appealing plots.

## Next Steps

- Explore additional `ggplot2` functionalities such as faceting and interactive plots.
- Use different datasets to practice visualization techniques.

This concludes our code-through tutorial on data visualization using `ggplot2`! 🎉
